Product graphics

### 2010-05-28 product-graphics.tex
Proof reading

@@

To construct plots for higher-dimensional tables, we need a way to decompose them into 1d and 2d components. To do we use some very basic properties of probability.  For example, let $f(x, y, z)$ be a 3d pmf. We can write this pmf as product of marginals and conditionals:

@@

\noindent Each of these decompositions is made up of products of one or two-d pmfs that can visualised used one of the graphical primitives described previously. Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.

@@

\subsection{Examples}
\label{sub:examples}

Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.

@@

Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.  Here we can see people tend to be happier the healthier and richer they are.

@@

\subsection{Existing plot types}

@@

% Other papers to cite:
%
%

\begin{itemize}
  \item {\bf Bar chart} (1d). 1 hbar.
  \item {\bf Column chart} (1d). 1 vbar.
  \item {\bf Spineplot} (1d). 1 vspine.
  \item {\bf Fluctuation} diagram (2d): fluct.
\end{itemize}

@@

\begin{itemize}
  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.
  \item {\bf Nested} bar chart (2d).  2 hbars. \citep{peltier:2009}

@@

  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}
  \item {\bf Generalised treemaps} (nd): any plot ending with a tile \citep{vliegen:2006}.

@@

Trellis graphics \citep{becker:1996}, as known as lattice or facetted or
conditioned graphics are another related display. They uses categorical
variables to generate multiple panels, each containing a plot of the subset of
the data. Trellised area plots also fall into our framework and can be
created by conditioning on the trellising variables.

>>>>

% See "Hierarchical Pixel Bar Charts" for constraints

@@

To construct plots of higher-dimensional tables, we need a way to decompose them into 1d and 2d components. To do so we use some very basic properties of probability. For example, let $f(x, y, z)$ be a 3d pmf. We can write this pmf as product of 1d and 2d marginals and conditionals:

@@

\noindent Since each component of the product is only 1 or 2d, we can use one of the primitives described above. For example, Figure \ref{fig:recursive} shows two ways to display the joint distribution of sex and happiness as broken down into the marginal distribution of happiness, and the condition distribution of sex, given happiness.

% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.

@@

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.  Here we can see people tend to be happier the healthier and richer they are.

@@

\section{Existing plot types}

@@

\begin{itemize}
  \item {\bf Bar chart} (1d). 1 hbar.
  \item {\bf Column chart} (1d). 1 vbar.
  \item {\bf Spineplot} (1d). 1 spine.
  \item {\bf Fluctuation} diagram (2d): 1 fluct.
\end{itemize}

@@

\begin{itemize}
  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.

  \item {\bf Nested} bar chart (2d).  2 hbars. \citep{peltier:2009}

@@

  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1981,friendly:1994,hofmann:2003}

  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}

  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}

  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}

  \item {\bf Generalised treemaps} (nd): any plot ending with a tile \citep{vliegen:2006}.

@@

Trellis graphics \citep{becker:1996}, also known as lattice, facetted and conditioned graphics, are another related display. They use categorical variables to generate multiple panels, each containing a plot of the subset of the data. Trellised area plots also fall into our framework and can be created by conditioning on the trellising variables.
